---
title: "Finding a relationship bewteen mRNA and Protein Expression"
output: html_document
---
## Preparing Data:

### 1) Load Data
```{r}
geneUN <- read.csv("https://raw.githubusercontent.com/gcohenfr/Neonatal-Vaccination/main/data/geneUN.csv?token=AMFFFE75VWA2N4KGLN6HGOLAWFKRM")
proteinUN <- read.csv("https://raw.githubusercontent.com/gcohenfr/Neonatal-Vaccination/main/data/proteinUN.csv?token=AMFFFE7LAEM3CRZTDJO6PJLAWFKVU")
```


### 2) Replace NA values
https://statisticsglobe.com/replace-missing-values-by-column-mean-in-r
```{r, warning=FALSE, message=FALSE}
for(i in 1:ncol(geneUN)) {   
  geneUN[ , i][is.na(geneUN[ , i])] <- median(geneUN[ , i], na.rm = TRUE)
}

for(i in 1:ncol(proteinUN)) {   
  proteinUN[ , i][is.na(proteinUN[ , i])] <- median(proteinUN[ , i], na.rm = TRUE)
}
```


### 3) Logarithmic Scale for proteinUN
```{r}
proteinUN.log <- proteinUN
for(i in 2:ncol(proteinUN.log)) {  
 proteinUN.log[ ,i] <- log2(proteinUN.log[ ,i] + 0.0000000001)
}
```


### 4) Viewing expression levels on plot


__Gene expression levels:__

```{r echo=FALSE}
plot(geneUN[,i],ylab="gene expression level",type="l",xlab="tissue type index",ylim=c(0,0.02))
```

__Protein expression levels:__

```{r echo=FALSE}
plot(proteinUN[,i],ylab="protein expression level",type="l",xlab="tissue type index",ylim=c(0,0.08))
```

__Log scaled protein expression levels:__

```{r echo=FALSE}
plot(proteinUN.log[,i],ylab="log2(protein expression level)",type="l",xlab="tissue type index",ylim=c(-30,0.08))
```

## Merging Data

### 1) Rename column names 

https://stackoverflow.com/questions/44064983/whats-the-best-way-to-add-a-specific-string-to-all-column-names-in-a-dataframe/44065175
https://statisticsglobe.com/rename-column-name-in-r-data-frame/

```{r include=FALSE}
library(data.table)
setnames(geneUN,paste(colnames(geneUN), sep="", "_ge"))
print(geneUN)
#need to rename just column X_ge -> X
colnames(geneUN)[colnames(geneUN) == "X_ge"] <- "X"
# renaming column names for proteinUN.log
setnames(proteinUN.log,paste(colnames(proteinUN.log), sep="", "_pe"))
colnames(proteinUN.log)[colnames(proteinUN.log) == "X_pe"] <- "X"
print(proteinUN.log)
```


### 2) Join dataframes
```{r include=FALSE}
library(dplyr)
mergedData <- full_join(proteinUN.log, geneUN.log, by = intersect(colnames(geneUN.log), colnames(proteinUN.log)))
```

```{r echo=FALSE}
head(mergedData)
```


## Analyzing Relationship between variables
http://mason.gmu.edu/~afinn/html/teaching/courses/f03_comm250/fbk_chapters/14.pdf
https://www.datacamp.com/community/tutorials/linear-regression-R#what

### Using Linear Regression Model

```{r include=FALSE}
library(ggplot2)
lmExpression = lm(uterus_pe~uterus_ge, data = mergedData)   #create linear regression
summary(lmExpression) #review results
```

__Uterus tissue:__

```{r, echo=FALSE, message=FALSE}
  ggplot(data=mergedData,aes(uterus_ge, uterus_pe)) +
  coord_cartesian(xlim = c(0, 0.002)) + 
  coord_cartesian(ylim = c(-25, 0)) +
  geom_point() +
  geom_smooth(method='lm')  #Add a regression line
```

__Kidney tissue__

```{r, echo=FALSE, message=FALSE}
  ggplot(data=mergedData,aes(kidney_ge, kidney_pe)) +
  coord_cartesian(xlim = c(0, 0.005)) + 
  coord_cartesian(ylim = c(-25, 0)) + 
  geom_point() + 
  geom_smooth(method='lm')  #Add a regression line
  kidney_model <- lm(kidney_ge~kidney_pe, data = mergedData)
  summary(kidney_model)
```


### Finding Correlation Coefficients

__Pearson Coefficients:__
* assumes normal distribution
```{r, echo=FALSE, message=FALSE}
library(corrplot)
mergedData.pcor <- cor(mergedData[sapply(mergedData,is.numeric)], method = c("pearson"))
corrplot(mergedData.pcor, method = "square")
```

__Spearman Coefficients:__
* assumes monotonic relationship
* does not assume normal distribution
```{r, echo=FALSE}
mergedData.scor <- cor(mergedData[sapply(mergedData,is.numeric)], method = c("spearman"))
corrplot(mergedData.scor, method = "square")
```

```{r, echo=FALSE}
cor.test(mergedData$uterus_ge, mergedData$uterus_pe,  method = "spearman", exact = FALSE) 
cor.test(mergedData$kidney_ge, mergedData$kidney_pe,  method = "spearman", exact = FALSE) 
```

rho > 0 -> positive correlation

